Update Meeting: 15-10-2021

### 2021-10-19 analysis/update_meeting_15_10_2021.Rmd
fix date in update meeting
add results to 15-10-21 meeting

options(scipen=999)
source("code/settings.R")
library(targets)
library(dplyr)
library(DT)
library(knitr)
library(ggplot2)
library(kableExtra)

@@

```{r load_targets, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}
tar_load(household_MR_summary)
tar_load(PCs_corr)
tar_load(PC_trait_corr)
tar_load(corr_impact_by_PCs)
```

@@

Our current filtering matches closely the v2 filtering. If we tried to mimic the v3 filtering, we could increase our sample size. What do we think about this discrepency? And more importantly, the decrease in power due to smaller sample size (about 25K less than the v3 parameters)?

@@

- Added back binary phenotypes. 
- Removed all dietary phenotypes (binary, ordinal and continuous).
- Removed some redundant phenotypes:
  - Left-side body traits (highly correlated with right-side) were removed
  - All 'qualifications' data (corresponding to field [6138](https://biobank.ctsu.ox.ac.uk/showcase/field.cgi?id=6138)) because data is roughly captured with continuous variables (such as years of education).
- See [phenotype selection](phenotpye_selection.html) document for complete summary of selection and remaining phenotypes.

@@

Restrict to only XY pairs that are have correlation < `0.8`. Correlations were calculated in the full biobank (unrelated, British, consenting etc.), not just within couples, in both sexes combined. Trait pairs and their correlations are summarized in the table below.

```{r corr_traits_summary, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

@@

In total, `#r length(which(household_MR_summary.DT$corr_traits > 0.8))` trait pairs were removed due to correlation > `0.8`.

**Note**: after filtering out pairs with high correlation, a bonferroni correction applied (p < 0.05 / [num remaining pairs after correaltion filter]) before examining paths from $X_i \rightarrow Y_p$ (i.e we only examined pairs that were significant along the diagonal).

@@

## MVMR

For each $X_i \rightarrow  Y_p$ MR, we included relevant covariates (Z's) in a MVMR framework in the $X_p \rightarrow Y_p$ MR. Z's were selected based on having BF significance in **both** of the MRs:

@@

## PC impact

@@

```{r PC_corr_couples, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

PCs_corr.DT <- PCs_corr %>% mutate(r = sqrt(as.numeric(r2))) 
DT::datatable(PCs_corr.DT, width = "100%")  %>% formatSignif(columns=c('r2', 'r'), digits=3)

@@

```{r PC_traits_corr, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

## merge the list of dataframes 
PCs_trait_corr.DT <- PC_trait_corr %>% mutate(r = sqrt(as.numeric(r2))) 
DT::datatable(PCs_corr.DT, width = "100%")  %>% formatSignif(columns=c('r2', 'r'), digits=3)

@@

```{r PC_trait_plot, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

plot <- ggplot2::ggplot(data = corr_impact_by_PCs, ggplot2::aes(x = trait_couple_corr,
                                                                y = corr_due_to_confounding, color = factor(overlap_above_50))) +
  geom_point(alpha = 3/4) +

  geom_smooth(mapping = aes(x = trait_couple_corr, y = corr_due_to_confounding), method = "lm", se=FALSE, formula = y~x+0, fullrange=TRUE, color = custom_col[4]) +
  geom_abline(slope=1, intercept=0, color = "black") +
  scale_color_manual(values = custom_col[c(1,2)], labels=c("overlap =< 50%", "overlap > 50%")) +
  theme_minimal() +
  ggplot2::labs(colour = legend_title, x = paste0("MR estimate\n(Y-index to Y-partner)"),
                y = paste("AM MR estimate adjusted for\nIV effect on relevant X"), title = "Comparison of adjusted and unadjusted AM estimated \n (sex-specific results)") + 
  theme(plot.title = element_text(hjust = 0.5))

@@

>>>>

options(scipen=999)
source("code/settings.R")
source("R/functions.R")
library(targets)
library(dplyr)
library(DT)
library(knitr)
library(ggplot2)
library(kableExtra)

@@

```{r load_targets, include = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}
tar_load(household_MR_summary)
tar_load(PCs_corr)
tar_load(PC_trait_corr)
tar_load(corr_impact_by_PCs)
tar_load(corr_impact_by_coords)
tar_load(proxyMR_MR_paths_summary_yiyp_adj)
tar_load(MV_z_corr_filter)
tar_load(traits_corr2_filled)
```

@@

Our current filtering matches closely the v2 filtering. If we tried to mimic the v3 filtering, we could increase our sample size. What do we think about this discrepancy? And more importantly, the decrease in power due to smaller sample size (about 25K less than the v3 parameters)? *This is just to keep in mind, if we want to increase our sample in the future if some results are nearly significant - could try adjusting our inclusion criteria.*

@@

- Added back binary phenotypes. 
- Removed all dietary phenotypes (binary, ordinal and continuous).
- Removed some redundant phenotypes:
  - Left-side body traits (highly correlated with right-side) were removed
  - All 'qualifications' data (corresponding to field [6138](https://biobank.ctsu.ox.ac.uk/showcase/field.cgi?id=6138)) because data is roughly captured with continuous variables (such as years of education).
- See [phenotype selection](phenotype_selection.html) document for complete summary of selection and remaining phenotypes.

@@

For the two trait MR, restrict to only XY pairs that are have correlation < `0.8`. Correlations were calculated in the full biobank (unrelated, British, consenting etc.), not just within couples, in both sexes combined. Trait pairs and their correlations are summarized in the table below.

```{r corr_traits_summary, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

@@

In total, `r length(which(household_MR_summary.DT$corr_traits > 0.8))` trait pairs were removed due to correlation > `0.8` from `r dim(household_MR_summary.DT)[1]` pairs.

**Note**: after filtering out pairs with high correlation, a Bonferroni correction applied (p < 0.05 / [num remaining pairs after correlation filter]) before examining paths from $X_i \rightarrow Y_p$ (i.e we only examined pairs that were significant along the diagonal).

@@

The results before and after adjustment are shown below:

```{r YiYp_adj, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE}
yiyp_adj.DT <- proxyMR_MR_paths_summary_yiyp_adj %>% dplyr::select(exposure_description, outcome_description, exposure_sex, yiyp_IVW_beta, yiyp_IVW_se, yiyp_IVW_pval, yiyp_IVW_beta_adj, yiyp_IVW_se_adj, yiyp_IVW_pval_adj)

format_round_cols <- colnames(dplyr::select_if(yiyp_adj.DT, is.numeric))
DT::datatable(yiyp_adj.DT, width = "100%") %>% formatRound(columns=format_round_cols, digits = 3)

## MVMR.

For each $X_i \rightarrow  Y_p$ MR, we included relevant co-variates (Z's) in a MVMR framework in the $X_p \rightarrow Y_p$ MR. Z's were selected based on having BF significance in **both** of the MRs:

@@

```{r MV_z, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE}

 z_list <- MV_z_corr_filter$MV_z_prune
average_num_z <- mean((sapply(z_list, length)))

On average, the number of Z's to include was: **`r average_num_z`**, which seems quite high.

See an example of Z's included for one trait pair below (first trait is exposure, second is outcome).

```{r MV_z_example, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE), message=FALSE}

i <- 1
exposure_description <- MV_z_corr_filter$exposure_description[i]
outcome_description <- MV_z_corr_filter$outcome_description[i]

exposure_description
outcome_description

z_list_ex <- tibble("z" = MV_z_corr_filter$MV_z_prune[i][[1]])

z_list_ex.DT <- replace_Neale_ID(z_list_ex, traits_corr2_filled, "z", "outcome_description")

DT::datatable(z_list_ex.DT, width = "100%")
```

## Impact of geography.

Wanted to assess impact of geography on couple trait correaltion. Investigated this by using both genetic PCs and birth coordinates.

### PC impact.

@@

```{r PC_corr_couples, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

PCs_corr.DT <- PCs_corr %>% mutate(r = sqrt(as.numeric(r2))) 
format_round_cols <- colnames(dplyr::select_if(PCs_corr.DT, is.numeric))
DT::datatable(PCs_corr.DT, width = "100%")  %>% formatRound(columns=format_round_cols, digits = 3)

@@

```{r PC_traits_corr, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

## merge the list of dataframes 
PCs_trait_corr.DT <- PC_trait_corr %>% bind_rows()
format_round_cols <- colnames(dplyr::select_if(PCs_trait_corr.DT, is.numeric))
DT::datatable(PCs_trait_corr.DT, width = "100%") %>% formatRound(columns=format_round_cols, digits = 3)

@@

```{r PC_trait_plot, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

plot <- ggplot2::ggplot(data = corr_impact_by_PCs, ggplot2::aes(x = trait_couple_corr,
                                                                y = corr_due_to_confounding)) +
  geom_point(alpha = 3/4) +

  geom_smooth(mapping = aes(x = trait_couple_corr, y = corr_due_to_confounding), method = "lm", se=FALSE, formula = y~x+0, fullrange=TRUE, color = custom_col[4]) +
  theme_minimal() +
  ggplot2::labs(x = paste0("Couple trait correlation"),
                y = paste("Couple correlation due to confounding"))

plot

@@

```{r PC_traits_confounding, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

## merge the list of dataframes 
corr_impact_by_PCs.DT <- replace_Neale_ID(corr_impact_by_PCs, traits_corr2_filled, "Neale_pheno_ID", "outcome_description")
format_round_cols <- colnames(dplyr::select_if(corr_impact_by_PCs.DT, is.numeric))
DT::datatable(corr_impact_by_PCs.DT, width = "100%")  %>% formatRound(columns=format_round_cols, digits = 3)

### Coordiante impact.

Performed the same analysis as above but replacing PCs with North and East birth co-ordinates (data field [129](https://biobank.ctsu.ox.ac.uk/crystal/field.cgi?id=129) and [130](https://biobank.ctsu.ox.ac.uk/crystal/field.cgi?id=130), respectively).

```{r coord_trait_plot, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

legend_title <- "Co-ordinate" 
plot <- ggplot2::ggplot(data = corr_impact_by_coords, ggplot2::aes(x = trait_couple_corr,
                                                                y = corr_due_to_confounding, color = factor(coordinate))) +
  geom_point(alpha = 3/4) +

  geom_smooth(mapping = aes(x = trait_couple_corr, y = corr_due_to_confounding), method = "lm", se=FALSE, formula = y~x+0, fullrange=TRUE, color = custom_col[4]) +
  scale_color_manual(values = custom_col[c(1,2)], labels=c("North co-ordinate", "East co-ordinate")) +
  theme_minimal() +
  ggplot2::labs(x = paste0("Couple trait correlation"),
                y = paste("Couple correlation due to confounding"))

plot

```{r coordin_traits_confounding, echo = FALSE, eval = ifelse(Sys.info()["sysname"]=="Linux", TRUE, FALSE)}

## merge the list of dataframes 
corr_impact_by_coords.DT <- replace_Neale_ID(corr_impact_by_coords, traits_corr2_filled, "Neale_pheno_ID", "outcome_description")
format_round_cols <- colnames(dplyr::select_if(corr_impact_by_coords.DT, is.numeric))
DT::datatable(corr_impact_by_coords.DT, width = "100%")  %>% formatRound(columns=format_round_cols, digits = 3)
